---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  fig.path = "man/figures/README-",
  dev.args = list(png = list(type = "cairo")),
  fig.retina = 2
)
```

# ggblend

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/ggblend)](https://CRAN.R-project.org/package=ggblend)
[![Codecov test coverage](https://codecov.io/gh/mjskay/ggblend/branch/main/graph/badge.svg)](https://app.codecov.io/gh/mjskay/ggblend?branch=main)
[![R-CMD-check](https://github.com/mjskay/ggblend/workflows/R-CMD-check/badge.svg)](https://github.com/mjskay/ggblend/actions)
<!-- badges: end -->

*ggblend* adds support for blend modes (like `"multiply"`, `"overlay"`, etc)
to *ggplot2*. It requires R >= 4.2, as blending and compositing support was added
in that version of R.

## Installation

You can install the development version of *ggblend* using:

```r
remotes::install_github("mjskay/ggblend")
```

## Blending within one geometry

We'll construct a simple dataset with two semi-overlapping point clouds. We'll
have two versions of the dataset: one with all the `"a"` points listed first,
and one with all the `"b"` points listed first.

```{r data}
library(ggplot2)
library(ggblend)
theme_set(theme_light())

set.seed(1234)
df_a = data.frame(x = rnorm(500, 0), y = rnorm(500, 1), set = "a")
df_b = data.frame(x = rnorm(500, 1), y = rnorm(500, 2), set = "b")

df_ab = rbind(df_a, df_b) |>
  transform(order = "draw a then b")

df_ba = rbind(df_b, df_a) |>
  transform(order = "draw b then a")

df = rbind(df_ab, df_ba)
```

A typical scatterplot of such data suffers from the problem that how many 
points appear to be in each group depends on the drawing order (*a then b* 
versus *b then a*):

```{r scatter_noblend}
df |>
  ggplot(aes(x, y, color = set)) +
  geom_point(size = 3) +
  scale_color_brewer(palette = "Set2") +
  facet_grid(~ order) +
  ggtitle("Scatterplot without blending: draw order matters")
```

A *commutative* blend mode, like `"multiply"` or `"darken"`, is one potential
solution that does not depend on drawing order. We can wrap `geom_point()` in
a call to `blend()` to achieve this, using something like `blend(geom_point(...), "darken")`
or equivalently `geom_point(...) |> blend("darken")`:

```{r scatter_blend}
df |>
  ggplot(aes(x, y, color = set)) +
  geom_point(size = 3) |> blend("darken") +
  scale_color_brewer(palette = "Set2") +
  facet_grid(~ order) +
  ggtitle("Scatterplot with blend('darken'): draw order does not matter")
```

Now the output is identical no matter the draw order.

## Alpha blending

One disadvantage of the `"darken"` blend mode is that it doesn't show density within
each group as well. We could use `"multiply"` instead, though then values can get quite dark:

```{r scatter_multiply}
df |>
  ggplot(aes(x, y, color = set)) +
  geom_point(size = 3) |> blend("multiply") +
  scale_color_brewer(palette = "Set2") +
  facet_grid(~ order) +
  ggtitle("Scatterplot with blend('multiply'): a little too dark")
```

Let's try combining two blend modes to address this: we'll use a `"lighten"`
blend mode (which is also commutative, but which makes the overlapping regions
lighter), and then draw the `"multiply"`-blended version on top at an `alpha`
of less than 1:

```{r scatter_lighten_multiply}
df |>
  ggplot(aes(x, y, color = set)) +
  geom_point(size = 3) |> blend("lighten") +
  geom_point(size = 3) |> blend("multiply", alpha = 0.65) +
  scale_color_brewer(palette = "Set2") +
  facet_grid(~ order) +
  ggtitle("Scatterplot with blend('lighten') and blend(multiply, alpha = 0.65)")
```

Now it's a little easier to see both overlap and density, and the output remains independent of draw order.

## Blending multiple geometries

We can also blend geometries together by passing a list of geometries to `blend()`.
These lists can include already-blended geometries:

```{r scatter_blend_geom}
df |>
  ggplot(aes(x, y, color = set)) +
  list(
    geom_point(size = 3) |> blend("darken"),
    geom_vline(xintercept = 0, color = "gray75", size = 1.5),
    geom_hline(yintercept = 0, color = "gray75", size = 1.5)
  ) |> blend("hard.light") +
  scale_color_brewer(palette = "Set2") +
  facet_grid(~ order) +
  labs(title = "Blending multiple geometries together")
```

## Blending groups within one geometry

Sometimes it's useful to have finer-grained control of blending within a given
geometry. Here, we'll show how to use `blend()` with `stat_lineribbon()`
from [ggdist](https://mjskay.github.io/ggdist/)
to create overlapping gradient ribbons depicting uncertainty.

We'll fit a model:

```{r m_mpg}
m_mpg = lm(mpg ~ hp * cyl, data = mtcars)
```

And generate some confidence distributions for the mean using [distributional](https://pkg.mitchelloharawild.com/distributional/):

```{r lineribbon}
grid = unique(mtcars[ c("cyl", "hp")])

predictions = grid |>
  cbind(predict(m_mpg, newdata = grid, se.fit = TRUE)) |>
  transform(mu_hat = distributional::dist_student_t(df = df, mu = fit, sigma = se.fit))

predictions
```

A basic plot based on examples in `vignette("freq-uncertainty-vis", package = "ggdist")` and
`vignette("lineribbon", package = "ggdist")` may have issues when lineribbons overlap:

```{r lineribbon_noblend}
predictions |>
  ggplot(aes(x = hp, fill = ordered(cyl), color = ordered(cyl))) +
  ggdist::stat_lineribbon(
    aes(ydist = mu_hat, fill_ramp = stat(.width)),
    .width = ppoints(40)
  ) +
  geom_point(aes(y = mpg), data = mtcars) +
  scale_fill_brewer(palette = "Set2") +
  scale_color_brewer(palette = "Dark2") +
  ggdist::scale_fill_ramp_continuous(range = c(1, 0)) +
  labs(
    title = "Overlapping lineribbons without blending", 
    color = "cyl", fill = "cyl", y = "mpg"
  ) +
  ggdist::theme_ggdist()
```

Notice the overlap of the orange (`cyl = 6`) and purple (`cyl = 8`) lines.

If we add a `blend_group = cyl` aesthetic mapping, we can blend the geometries
for the different levels of `cyl` together with a `blend()` call around
`ggdist::stat_lineribbon()`:

```{r lineribbon_blend}
predictions |>
  ggplot(aes(x = hp, fill = ordered(cyl), color = ordered(cyl), blend_group = cyl)) +
  ggdist::stat_lineribbon(
    aes(ydist = mu_hat, fill_ramp = stat(.width)),
    .width = ppoints(40)
  ) |> blend("multiply") +
  geom_point(aes(y = mpg), data = mtcars) +
  scale_fill_brewer(palette = "Set2") +
  scale_color_brewer(palette = "Dark2") +
  ggdist::scale_fill_ramp_continuous(range = c(1, 0)) +
  labs(
    title = "Overlapping lineribbons with blend('multiply')", 
    color = "cyl", fill = "cyl", y = "mpg"
  ) +
  ggdist::theme_ggdist()
```

Now the overlapping ribbons are blended together.
